---
title: "Serological Assay Quality Control Report"
output: 
  html_document:
    theme: cosmo
    code_folding: hide
params:
  raw_data_filename: NA
  plate_layout_filename: NA
  experiment_name: NA
  experiment_notes: NA
  date: NA
  stdcurve_plot: NA
  plateqc_plot: NA
  blanks_plot: NA
  model_results: NA
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, warning = FALSE, message = FALSE, tidy = TRUE)
require(tidyverse)
require(readxl)
require(drc)
require(janitor)
require(plyr)
require(here)
```

***
### QC Report for experiment: `r params$experiment_name` analyzed on `r params$date` {.tabset}
***

#### Plot of standard curves
```{r}
params$stdcurve_plot
```

#### Plot of model results
```{r}
model_results <- params$model_results

plots_model <- lapply(seq_along(model_results), function(x){
    ggplot(data = model_results[[x]], 
           aes(x = dilution, y = `1`)) + 
      geom_line() +
      scale_x_log10(breaks = c(1e-5, 1e-4, 1e-3, 1e-2, 0.03),
                    labels = c("0.00001", "0.0001", "0.001", "0.01", "0.03")) +
      labs(x = "antibody dilution",
           y = "standard curve",
           title = names(model_results[x])) +
      theme_bw()
  })
  
gridExtra::grid.arrange(grobs = plots_model, nrow = 3)
```

#### Plot of bead counts
Any wells with <15 beads/well are indicated in red, and should be double-checked manually.
```{r}
params$plateqc_plot
```

#### Plot of blank samples
Blank samples should record MFI<50 for each protein. If any proteins are above the dashed line, they should be double-checked manually.
```{r}
params$blanks_plot
```

#### Experiment info

**Experiment notes:** `r params$experiment_notes`

**Raw data filename:** `r params$raw_data_filename`

**Plate layout filename:** `r params$plate_layout_filename`